A Formal Model of Checked C

### 2023-01-08 sandp23/background.tex
update.

\subsection{\checkedc}
% 
\checkedc{} development began in 2015 by Microsoft Research, but it was forked
in late 2021 and is now actively managed by the Secure Software
Development Project (SSDP). Details can be found in a prior
overview~\cite{Elliott2018} or the formalism~\cite{li22checkedc}.
%

\noindent
\myparagraph{Checked Pointer Types}
\checkedc{} introduces three varieties of \emph{checked pointer}:
\begin{itemize}
\item \code{ptr<}$T$\code{>} ($\ptr$) types a pointer that is either null or
  points to a single object of type $T$.
\item \code{array_ptr<}$T$\code{>} ($\arr$) types a pointer that is either null
  or points to an array of $T$ objects. The array width is defined
  by a \emph{bounds} expression, discussed below.
\item \code{nt_array_ptr<}$T$\code{>} ($\ntarr$) is like
  \code{array_ptr<}$T$\code{>} except that the bounds expression
  defines the \emph{minimum} array width---additional objects may
  be available past the upper bound, up to a null terminator.
\end{itemize}
Both $\arr$ and $\ntarr$ pointers have an associated bounds which defines the
range of memory referenced by the pointer.
The three different ways to declare bounds and the corresponding memory range is:
\begin{footnotesize}
\begin{tabular}{ll}
\arrT{|$T$|} \inlinecode{p: count(|$n$|)}
  &
$[\inlinecode{p}, \inlinecode{p}+\inlinecode{sizeof}(T) \times n) $ \\
%$[\inlinecode{p}, )$ \\
\arrT{|$T$|} \inlinecode{p: byte_count(|$b$|)}

@@

Oftentimes, inserted checks can be optimized away by LLVM.
Consider the above code to be enclosed in another condition, such as,~\inlinecode{if (n<u) if (n>0) a[n-1] =}.
In such cases, the inserted check can be
removed as the outer condition $\inlinecode{n<u}$ already ensures that the
dereference is within bounds.
Dual~\etal~\cite{duanrefactoring} found essentially no
overhead when running Checked C-converted portions of the FreeBSD kernel.

@@

\myparagraph{Converting C to \checkedc}
%
The safety guarantees of Checked C come with certain restrictions. For instance,
as shown below, Checked C programs cannot use address-taken variables in a
bounds expression as the bounds relations may not hold because of possible
modifications through pointers.
% 
\begin{minted}[xleftmargin=30pt, mathescape, escapeinside=||, fontsize=\footnotesize]{c}
...
array_ptr<int> p : count (n) = NULL;
|\textcolor{red}{\faTimes}|..,&n,.
\end{minted}
% 
Consequently, converting existing C programs to Checked C might require
refactoring,~\eg eliminate~\inlinecode{&n} from the program above without
changing its functionality.
% 
This might require considerable effort~\cite{duanrefactoring} depending on the
program's complexity.
% 
Recently, Machiry~\etal developed~\threec~\cite{machiry2022c} that tries to
automatically convert a program to Checked C by adding appropriate pointer
annotations.
However, as described in \threec, completely automated conversion
is \emph{infeasible}, and it requires the developer to convert some code regions
manually.

@@

\liyi{I do not think you need three sub-sections for these topics. They can be in the same place. }
\subsection{Program Partitioning}
\label{subsec:background:programpart}
Program partitioning~\cite{rul2009towards} is a well-known technique to divide a program into multiple isolated parts or partitions.
There has been considerable work~\cite{tan2017principles, brumley2004privtrans, bittau2008wedge, lind2017glamdring, liu2017ptrsplit} in the area.
Most of these techniques are~\emph{data-centric}~\cite{lind2017glamdring, liu2017ptrsplit}, wherein program data drives the partitioning.
Specifically, given sensitive data in a program, the goal is to partition functions into two parts or partitions based on whether a function can access the sensitive data.
The performance overhead of these approaches is dominated by marshaling costs and depends on the usage of sensitive data.
The overhead of state-of-the-art approaches, such as Glamdring~\cite{lind2017glamdring} and  PtrSplit~
\cite{liu2017ptrsplit}, is prohibitive and varies from 37\%-163\%.

@@

\myparagraph{Converting C to \checkedc}
%
The safety guarantees of Checked C come with certain restrictions. For instance,
as shown below, Checked C programs cannot use address-taken variables in a
bounds expression as the bounds relations may not hold because of possible
modifications through pointers.
% 
\begin{minted}[xleftmargin=30pt, mathescape, escapeinside=||, fontsize=\footnotesize]{c}
...
array_ptr<int> p : count (n) = NULL;
|\textcolor{red}{\faTimes}|..,&n,.
\end{minted}
% 
Consequently, converting existing C programs to Checked C might require
refactoring, e.g., eliminate~\inlinecode{&n} from the program above without
changing its functionality.
% 
This might require considerable effort~\cite{duanrefactoring} depending on the
program's complexity.
% 
Recently, Machiry et al. developed~\threec~\cite{machiry2022c} that tries to
automatically convert a program to Checked C by adding appropriate pointer
annotations.  However, as described in \threec, completely automated conversion
is \emph{infeasible}, and it requires the developer to convert some code regions
manually.  
% 
Although the backward compatibility of Checked C helps a partially annotated
program to enjoy spatial memory safety on those regions using only Checked
pointers (i.e., checked, or safe regions), \mz{then what?}.

>>>>

\subsection{\checkedc}
\label{subsec:checkedc}
% 
\checkedc{} development began in 2015 by Microsoft Research, but it was forked
in late 2021 and is now actively managed by the Secure Software
Development Project (SSDP). Details can be found in a prior
overview~\cite{Elliott2018} or the formalism~\cite{li22checkedc}.
%

\liyi{Do we need to spend more than half page introducing checked-c? OR just introduce it briefly like what we had in the formalism paper? }

\noindent
\myparagraph{Checked Pointer Types}
\checkedc{} introduces three varieties of \emph{checked pointer}:
\begin{itemize}
\item \code{_Ptr<}$T$\code{>} ($\ptr$) types a pointer that is either null or
  points to a single object of type $T$.
\item \code{_Array_ptr<}$T$\code{>} ($\arr$) types a pointer that is either null
  or points to an array of $T$ objects. The array width is defined
  by a \emph{bounds} expression, discussed below.
\item \code{_NT_Array_ptr<}$T$\code{>} ($\ntarr$) is like
  \code{_Array_ptr<}$T$\code{>} except that the bounds expression
  defines the \emph{minimum} array width---additional objects may
  be available past the upper bound, up to a null terminator.
\end{itemize}
Both $\arr$ and $\ntarr$ pointers have an associated bounds which defines the
range of memory referenced by the pointer.
The three different ways to declare bounds and the corresponding memory range is:
\begin{footnotesize}
\begin{tabular}{ll}
\arrT{|$T$|} \inlinecode{p: count(|$n$|)}
  &
$[\inlinecode{p}, \inlinecode{p}+\inlinecode{sizeof}(T) \times n) $ \\
%$[\inlinecode{p}, )$ \\
\arrT{|$T$|} \inlinecode{p: byte_count(|$b$|)}

@@

\liyi{why we need the following? I suggest we can cut. }
Oftentimes, inserted checks can be optimized away by LLVM.
Consider the above code to be enclosed in another condition, such as,~\inlinecode{if (n<u) if (n>0) a[n-1] =}.
In such cases, the inserted check can be
removed as the outer condition $\inlinecode{n<u}$ already ensures that the
dereference is within bounds.
Dual~\etal~\cite{duanrefactoring} found essentially no
overhead when running Checked C-converted portions of the FreeBSD kernel.

@@

\myparagraph{Converting C to \checkedc}
%
The safety guarantees of Checked C come with certain restrictions. For instance,
as shown below, Checked C programs cannot use address-taken variables in a
bounds expression as the bounds relations may not hold because of possible
modifications through pointers.
% 
\begin{minted}[xleftmargin=30pt, mathescape, escapeinside=||, fontsize=\footnotesize]{c}
...
_Array_ptr<int> p : count (n) = NULL;
|\textcolor{red}{\faTimes}|..,&n,.
\end{minted}
% 
Consequently, converting existing C programs to Checked C might require
refactoring,~\eg eliminate~\inlinecode{&n} from the program above without
changing its functionality.
% 
This might require considerable effort~\cite{duanrefactoring} depending on the
program's complexity.
% 
Recently, Machiry~\etal developed~\threec~\cite{machiry2022c} that tries to
automatically convert a program to Checked C by adding appropriate pointer
annotations.
However, as described in \threec, completely automated conversion
is \emph{infeasible}, and it requires the developer to convert some code regions
manually.

@@

\subsection{Program Partitioning}
\label{subsec:background:programpart}
Program partitioning~\cite{rul2009towards} is a well-known technique to divide a program into multiple isolated parts or partitions.
There has been considerable work~\cite{tan2017principles, brumley2004privtrans, bittau2008wedge, lind2017glamdring, liu2017ptrsplit} in the area.
Most of these techniques are~\emph{data-centric}~\cite{lind2017glamdring, liu2017ptrsplit}, wherein program data drives the partitioning.
Specifically, given sensitive data in a program, the goal is to partition functions into two parts or partitions based on whether a function can access the sensitive data.
The performance overhead of these approaches is dominated by marshaling costs and depends on the usage of sensitive data.
The overhead of state-of-the-art approaches, such as Glamdring~\cite{lind2017glamdring} and  PtrSplit~
\cite{liu2017ptrsplit}, is prohibitive and varies from 37\%-163\%.

@@

\myparagraph{Converting C to \checkedc}
%
The safety guarantees of Checked C come with certain restrictions. For instance,
as shown below, Checked C programs cannot use address-taken variables in a
bounds expression as the bounds relations may not hold because of possible
modifications through pointers.
% 
\begin{minted}[xleftmargin=30pt, mathescape, escapeinside=||, fontsize=\footnotesize]{c}
...
_Array_ptr<int> p : count (n) = NULL;
|\textcolor{red}{\faTimes}|..,&n,.
\end{minted}
% 
Consequently, converting existing C programs to Checked C might require
refactoring, e.g., eliminate~\inlinecode{&n} from the program above without
changing its functionality.
% 
This might require considerable effort~\cite{duanrefactoring} depending on the
program's complexity.
% 
Recently, Machiry et al. developed~\threec~\cite{machiry2022c} that tries to
automatically convert a program to Checked C by adding appropriate pointer
annotations.  However, as described in \threec, completely automated conversion
is \emph{infeasible}, and it requires the developer to convert some code regions
manually.  
% 
Although the backward compatibility of Checked C helps a partially annotated
program to enjoy spatial memory safety on those regions using only Checked
pointers (i.e., checked, or safe regions), \mz{then what?}.
